---
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

title: "R Notebook" output: html_notebook

#harlingen1889_bevolking huisgezinnen = 2133 hoofd_huisgezinnen =
idem_huisgezinnen afzonderlijk_levenden = 221 overige_leden = 2310
tax_records = 978

# libraries

```{r}
library("data.table")
library("arrow")
library("ineq")     # inequality measures and example data
library("EnvStats") # for estimation of censored distributions
source("impute.R") 
```

# data tax_record

```{r}
tax = read_parquet("/Users/werk/Documents/git_hipnl/data/tax_records/tax_records_latest.parquet")
setDT(tax)

tax = tax[place == "Harlingen" & year == 1889]

y_trunc = na.omit(tax$income_taxable)
n_tax_records = length(y_trunc)
n_tax_records
```

```{r}
#harlingen1889_bevolking
huisgezinnen = 2133 
hoofd_huisgezinnen = huisgezinnen
afzonderlijk_levenden = 221
overige_leden = 2310

```

```{r}
cat1 = huisgezinnen + afzonderlijk_levenden
cat2 = afzonderlijk_levenden + (huisgezinnen*2) + overige_leden
cat3 = afzonderlijk_levenden + huisgezinnen + overige_leden
#proportioneel = x*cat1 + x*cat2 + x*cat3
```

```{r}
categories = c(cat1 = cat1, cat2 = cat2, cat3 = cat3)
n_missing_by_cat = categories - n_tax_records
n_missing_by_cat
```

```{r}
# naive estimates
naive_estimates = c(
  gini = ineq::Gini(y_trunc),
  mean_log = mean(log(y_trunc), na.rm = TRUE),
  sd_log = sd(log(y_trunc), na.rm = TRUE)
)
naive_estimates
```

# mean; sd; zonder imputaties lnorm

```{r}
sapply(n_missing_by_cat, estimate_lnorm, y = y_trunc)


```

# gini_sd

```{r}
# impute_from_sd() uses the sd from the lognormal to estimate the Gini
toplot_from_sd = sapply(n_missing_by_cat, impute_from_sd, y = y_trunc)
toplot_from_sd
```

```{r}
parameters = estimate_lnorm(y_trunc, n_missing_by_cat[1])
parameters
```

```{r}
c(
        y_trunc, 
        rep(min(y_trunc, na.rm = TRUE), n_missing_by_cat[1])
    )
```

```{r}
imputations = rnormTrunc(
        n = n_missing_by_cat[1], 
        mean = parameters$log_mean, 
        sd = parameters$log_sd, 
        min = -Inf,
        max = min(log(y_trunc)))
sort(exp(imputations))
```

```{r}
pnorm(parameters$log_sd / sqrt(2))
```

```{r}
out = c(y_trunc, exp(imputations))
out_dens <- density(out) # returns the density data 
plot(out_dens, )
# out < 500
```

# gini_lnorm

```{r}
# list of samples
resampled = lapply(n_missing_by_cat, sample_from_lognormal, y = y_trunc)
# gini on each sample in the list
toplot_resampled = sapply(resampled, ineq::Gini)
toplot_resampled

```

```{r}
toplot_resampled - toplot_from_sd
```

```{r}
barplot(toplot_resampled, main = "gini_resampled_lnorm", ylim = c(0,1))
barplot(toplot_from_sd, main = "gini_sd_method", ylim = c(0,1))
```

# proportioneel

```{r}
categories_proportions = c(cat1 = 0.83, cat2 = 0.04, cat3 = 0.13)
n_categories_proportional = categories_proportions * categories
n_categories_proportional

n_proportional_missing = sum(n_categories_proportional)
n_proportional_missing






```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or
by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output
will be saved alongside it (click the *Preview* button or press
*Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the
editor. Consequently, unlike *Knit*, *Preview* does not run any R code
chunks. Instead, the output of the chunk when it was last run in the
editor is displayed.
